\begin{center}
{\bf Author's Biographical Sketch}
\end{center}

Dr. Norm Matloff is a professor of computer science at the University of
California at Davis, and was formerly a professor of statistics at that
university. He is a former database software developer in Silicon
Valley, and has been a statistical consultant for firms such as the
Kaiser Permanente Health Plan.

Dr. Matloff was born in Los Angeles, and grew up in East Los Angeles and
the San Gabriel Valley. He has a PhD in pure mathematics from UCLA,
specializing in probability theory and statistics.  He has published
numerous papers in computer science and statistics, with current
research interests in parallel processing, statistical computing,
and regression methodology.

Prof. Matloff is a former appointed member of IFIP Working Group 11.3,
an international committee concerned with database software security,
established under UNESCO.  He was a founding member of the UC Davis
Department of Statistics, and participated in the formation of the UCD
Computer Science Department as well.  He is a recipient of the
campuswide Distinguished Teaching Award and Distinguished Public Service
Award at UC Davis.

Dr. Matloff is the author of two published textbooks, and of a number of
widely-used Web tutorials on computer topics, such as the Linux
operating system and the Python programming language.  He and Dr. Peter
Salzman are authors of {\it The Art of Debugging with GDB, DDD, and
Eclipse}. Prof.  Matloff's book on the R programming language, {\it The
Art of R Programming}, was published in 2011.  His book, {\it Parallel
Computation for Data Science}, came out in 2015.  His current book
project, {\it From Linear Models to Machine Learning: Predictive
Insights through R}, will be published in 2016.  He is also the author
of several open-source textbooks, including {\it From Algorithms to
Z-Scores: Probabilistic and Statistical Modeling in Computer Science}
(\url{http://heather.cs.ucdavis.edu/probstatbook}), and {\it Programming
on Parallel Machines}
(\url{http://heather.cs.ucdavis.edu/~matloff/ParProcBook.pdf}).
