\chapter{Transform Methods}
\label{chap:xform}

We often use the idea of \textbf{transform} functions.  For example, you
may have seen \textbf{Laplace transforms} in a math or engineering
course.  The functions we will see here differ from this by just
a change of variable. 

Though in the form used here they involve only univariate distributions,
their applications are often multivariate, as will be the case here.

\section{Generating Functions}

Let's start with the \textbf{generating function}.
For any nonnegative-integer valued
random variable V, its generating function is defined by

\begin{equation}
\label{genftn}
g_V(s)=E(s^{V}) = \sum_{i=0}^\infty s^i p_V(i), ~ 0 \leq s \leq 1
\end{equation}

For instance, suppose N has a geometric distribution with parameter p,
so that $p_N(i) = (1-p) p^{i-1}$, i = 1,2,...  Then

\begin{equation}
g_N(s) = \sum_{i=1}^\infty s^i \cdot (1-p) p^{i-1} 
= \frac{1-p}{p} \sum_{i=1}^\infty s^i \cdot p^i 
= \frac{1-p}{p} \frac{ps}{1-ps}
= \frac{(1-p)s}{1-ps}
\end{equation}

Why restrict s to the interval [0,1]?  The answer is that for $s > 1$
the series in (\ref{genftn}) may not converge.  for $0 \leq s \leq 1$,
the series does converge.  To see this, note that if s = 1, we just get
the sum of all probabilities, which is 1.0.  If a nonnegative s is less
than 1, then $s^i$ will also be less than 1, so we still have
convergence.

One use of the generating function is, as its name implies, to generate
the probabilities of values for the random variable in question.  In
other words, if you have the generating function but not the
probabilities, you can obtain the probabilities from the function.
Here's why:  For clarify, write (\ref{genftn}) as

\begin{equation}
g_V(s) = P(V=0) + s P(V=1) + s^2 P(V=2) + ...  
\end{equation}

From this we see that 

\begin{equation}
g_V(0) = P(V = 0)
\end{equation}

So, we can obtain P(V = 0) from the generating function.  Now
differentiating (\ref{genftn}) with respect to s, we have

\begin{eqnarray}
g'_V(s) &=&
\frac{d}{ds} \left [ P(V=0) + s P(V=1) + s^2 P(V=2) + ...  \right ]
\nonumber \\
&=&
P(V=1) + 2s P(V=2) + ...  
\end{eqnarray}

So, we can obtain P(V = 2) from $g'_V(0)$, and in a similar manner can
calculate the other probabilities from the higher derivatives.

\section{Moment Generating Functions}

The generating function is handy, but it is limited to discrete random
variables.  More generally, we can use the {\bf moment generating
function}, defined for any random variable X as

\begin{equation}
\label{momgen}
m_X(t) = E[e^{tX}] 
\end{equation}

for any t for which the expected value exists.

That last restriction is anathema to mathematicians, so they use the
characteristic function, 

\begin{equation}
\phi_X(t) = E[e^{itX}] 
\end{equation}

which exists for any t.  However, it makes use of pesky complex numbers,
so we'll stay clear of it here.

Differentiating (\ref{momgen}) with respect to t, we have

\begin{equation}
m_X'(t) = E[X e^{tX}] 
\end{equation}

We see then that

\begin{equation}
m_X'(0) = EX
\end{equation}

So, if we just know the moment-generating function of X, we can obtain
EX from it.  Also,

\begin{equation}
m_X''(t) = E(X^2 e^{tX})
\end{equation}

so

\begin{equation}
m_X''(0) = E(X^2)
\end{equation}

In this manner, we can for various k obtain $E(X^k)$, the {\bf k$th$
moment} of X, hence the name.

\section{Transforms of Sums of Independent Random Variables}
\label{tranfssums}

Suppose X and Y are independent and their moment generating functions
are defined.  Let Z = X+Y.  then

\begin{equation}
m_Z(t) = E[e^{t(X+Y)}] = E[e^{tX} e^{tY}] = E(e^{tX}) \cdot E(e^{tY}) =
m_X(t) m_Y(t)
\end{equation}

In other words, the mgf of the sum is the product of the mgfs!  This is
true for other transforms, by the same reasoning.

Similarly, it's clear that the mgf of a sum of three independent
variables is again the product of their mgfs, and so on.

\section{Example:  Network Packets}

As an example, suppose say the number of packets N received on a network
link in a given time period has a Poisson distribution with mean $\mu$,
i.e.

\begin{equation}
P(N = k) = \frac{e^{-\mu} \mu^k}{k!}, k = 0,1,2,3,...  
\end{equation}

\subsection{Poisson Generating Function}

Let's first find its generating function.

\begin{equation}
g_N(t) = \sum_{k=0}^{\infty} t^k  \frac{e^{-\mu} \mu^k}{k!} 
= e^{-\mu} \sum_{k=0}^{\infty} \frac{(\mu t)^k}{k!}
= e^{-\mu+\mu t}
\end{equation}

where we made use of the Taylor series from calculus,

\begin{equation}
e^u =  \sum_{k=0}^{\infty} u^k/k!
\end{equation}

\subsection{Sums of Independent Poisson Random Variables Are Poisson
Distributed}
\label{sumpoi}

Supposed packets come in to a network node from two independent links,
with counts $N_1$ and $N_2$, Poisson distributed with means $\mu_1$ and
$\mu_2$.  Let's find the distribution of $N = N_1+N_2$, using a
transform approach.

From Section \ref{tranfssums}:

\begin{equation}
\label{sumpoieqn}
g_N(t) = g_{N_1}(t) g_{N_2}(t) = e^{-\nu + \nu t}
\end{equation}

where $\nu = \mu_1 + \mu_2$.

But the last expression in (\ref{sumpoieqn}) is the generating function for
a Poisson distribution too!  And since there is a one-to-one
correspondence between distributions and transforms, we can conclude
that N has a Poisson distribution with parameter $\nu$.  We of
course knew that N would have mean $\nu$ but did not know that N would
have a Poisson distribution.

So:  A sum of two independent Poisson variables itself has a Poisson
distribution.  By induction, this is also true for sums of k independent
Poisson variables.

% \section{Random Number of Bits in Packets on One Link}
% \label{randomn}
% 
% Consider just one of the two links now, and for convenience denote the
% number of packets on the link by N, and its mean as $\mu$.  Continue to
% assume that N has a Poisson distribution.
% 
% Let B denote the number of bits in a packet, with $B_1,...,B_N$ denoting
% the bit counts in the N packets.  We assume the $B_i$ are independent
% and identically distributed. The total number of bits received during
% that time period is
% 
% \begin{equation}
% T=B_1+...+B_N
% \end{equation}
% 
% Suppose the generating function of B is known to be h(s). Then what is
% the generating function of T?
% 
% \begin{eqnarray}
% g_T(s) & = & E(s^{T})\\
%  & = & E[E(s^{T}|N)]\\
%  & = & E[E(s^{B_1+...+B_N}|N)]\\
%  & = & E[E(s^{B_1}|N)...E(s^{B_N}|N)]\\
%  & = & E[h(s)^{N}]\\
%  & = & g_N[h(s)]\\
%  & = & e^{-\mu+\mu h(s)}
% \end{eqnarray}  
% 
% Here is how these steps were made: 
% 
% \begin{itemize}
% 
% \item From the first line to the second, we used
% the Theorem of Total Expectation. 
% 
% \item From the second to the third, we just used the definition of T.
% 
% \item From the third to the fourth lines, we have used algebra plus the
% fact that the expected value of a product of independent random
% variables is the product of their individual expected
% values.
% 
% \item From the fourth to the fifth, we used the definition of h(s).
% 
% \item From the fifth to the sixth, we used the definition of $g_N$.
% 
% \item From the sixth to the last we used the formula for the generating
% function for a Poisson distribution with mean $\mu$.
% 
% \end{itemize}
% 
% We can then get all the information about T we need from this formula, such
% as its mean, variance, probabilities and so on, as seen previously. 

\section{Other Uses of Transforms}

Transform techniques are used heavily in queuing analysis, including for
models of computer networks.  The techniques are also used extensively
in modeling of hardware and software reliability.

Transforms also play key roles in much of theoretical probability, the
Central Limit Theorems\footnote{The plural is used here because there
are many different versions, which for instance relax the condition that
the summands be independent and identically distributed.} being a good
example.  Here's an outline of the proof of the basic CLT, assuming the
notation of Section \ref{formalclt}:

First rewrite Z as

\begin{equation}
Z = \sum_{i=1}^n \frac{X_i - m}{v \sqrt{n}}
\end{equation}

Then work with the characteristic function of Z:

\begin{eqnarray}
c_Z(t) &=& E(e^{itZ}) ~~~~ \textrm{(def.)} \\ 
&=& \Pi_{i=1}^n E[e^{it(X_i-m)/(v\sqrt{n})}] ~~~~ \textrm{(indep.)} \\
&=&  \Pi_{i=1}^n E[e^{it(X_1-m)/(v\sqrt{n})}] ~~~~ \textrm{(ident. distr.)} \\
&=& [g(\frac{it}{\sqrt{n}})]^n
\label{gpower}
\end{eqnarray}

where g(s) is the characteristic function of $(X_1-m)/v$, i.e.

\begin{equation}
g(s) = E[e^{is \cdot \frac{X_1 - m}{v}}]
\end{equation}

Now expand (\ref{gpower}) in a Taylor series around 0, and use the fact
that $g'(0)$ is the expected value of $(X_1-m)/v$, which is 0:

\begin{eqnarray}
[g(\frac{t}{\sqrt{n}})]^n 
&=& \left [1 - \frac{t^2}{2n} + o(\frac{t^2}{n}) \right ]^n \\
&\rightarrow& e^{-t^2/2} \textrm{ as } n \rightarrow \infty
\label{ourcltgoal}
\end{eqnarray}

where we've also used the famous fact that $(1 - s/n)^n$ converges to
$e^{-s}$ as $n \rightarrow \infty$.

But (\ref{ourcltgoal}) is the density of N(0,1), so we have proved the
Central Limit Theorem.

\startproblemset

\oneproblem
Use transform methods to derive some properties of the Poisson family:

\begin{itemize}

\item [(a)] Show that for any Poisson random variable, its mean and
variance are equal.

\item [(b)] Suppose $X$ and $Y$ are independent random variables, each
having a Poisson distribution.  Show that $Z = X+Y$ again has a Poisson
distribution.

\end{itemize}

\oneproblem
In our ordinary coins which we use every day, each one has a slightly 
different probability of heads, which we'll call $H$.  Say $H$ has 
the distribution $N(0.5, 0.03^2)$.  We choose a coin from a batch at 
random, then toss it 10 times.  Let $N$ be the number of heads we get.  
Find $Var(N)$.


\oneproblem
Suppose the number N of bugs in a certain number of lines of code has a
Poisson distribution, with parameter L, where L varies from one
programmer to another. Show that Var(N) = EL + Var(L).

\oneproblem
Let X denote the number we obtain when we roll a single die
once.  Let $G_X(s)$ denote the generating function of X.

\begin{itemize}

\item [(a)] Find $G_X(s)$.

\item [(b)] Suppose we roll the die 5 times, and let T denote the
total number of dots we get from the 5 rolls.  Find $G_T(s)$.

\end{itemize}

% Plot of 2-D normal, adapted from R Graph Gallery,
% http://addictedtor.free.fr/graphiques/graphcode.php?graph=42:
% 
% mu1 <- 0  # EX1
% mu2 <- 0  # EX2
% s11 <- 10  # Var(X1) = 10
% s22 <- 10  # Var(X2) = 15
% x1 <- seq(-10,10,length=41) 
% x2 <- x1 
% 
% # bivariate normal density
% f <- function(x1,x2,rho){
%         term1 <- 1/(2*pi*sqrt(s11*s22*(1-rho^2)))                               
%         term2 <- -1/(2*(1-rho^2))                                               
%         term3 <- (x1-mu1)^2/s11                                                 
%         term4 <- (x2-mu2)^2/s22                                                 
%         term5 <- 2*rho*((x1-mu1)*(x2-mu2))/(sqrt(s11)*sqrt(s22))               
%         term1*exp(term2*(term3+term4-term5))                                    
% } 
% 
% plot2dnorm <- function(x1,x2,f,rho) {
%    z <- outer(x1,x2,f,rho)
%    persp(x1, x2, z,
%       col="lightgreen",                                                         
%       theta=30, phi=20,                                                         
%       r=50,                                                                     
%       d=0.1,                                                                    
%       expand=0.5,                                                               
%       ltheta=90, lphi=180,                                                      
%       shade=0.75,                                                               
%       ticktype="detailed",                                                      
%       nticks=5) 
% }
% 
% 
% 
% # simulation of dice game; NOT efficient R code
% 
% # game:  roll a die 50 times; each time, get $5 for a 1, $2 for a 2 or 3
% 
% # random multinomial generator (R has one too)
% rmn <- function(pcdf) {  # pcdf is the cdf vector of the multinomial
%    u <- runif(1)
%    ncats <- length(pcdf)
%    for (i in 1:(ncats-1))
%       if (u <= pcdf[i]) return(i)
%    return(ncats)
% }
% 
% # do the simulation
% nplays <- 50
% nreps <- 1000
% p1 <- 1/6
% p23 <- 3/6
% prest <- 1
% nums <- matrix(nrow=nreps,ncol=2)
% wins <- vector(length=nreps)
% for (rep in 1:nreps) {
%    totwin <- 0
%    numones <- 0
%    numtwosthrees <- 0
%    for (i in 1:nplays) {
%       r <- rmn(c(p1,p23,prest))
%       if (r == 1) numones <- numones + 1
%       else if (r == 2) numtwosthrees <- numtwosthrees + 1
%    }
%    nums[rep,] <- c(numones,numtwosthrees)
%    wins[rep] <- 5*numones + 2*numtwosthrees
% }
% 
% # analyze the data
% # find F(,)
% up1 <- 12
% up2 <- 16
% print(nrow(nums[nums[,1] <= up1 & nums[,2] <= up2,])/nreps)
% # winnings should be approx normal distributed
% hist(wins)
% # find P(win > $90)
% print(length(wins[wins>90])/nreps)
% 
% 
% # plotting the double integral
% plot(c(0,1),c(0,1),type="n",xlab="s",ylab="t")
% polygon(c(0,1,1,0),c(0,0,1,0),col="gray")
% polygon(c(1,1,0.5,1),c(0,1,0.5,0),density=5)
% lines(c(0.7,0.7),c(0.3,0.7),lwd=3)
% text(locator(1),"t=1-s")
% locator(1)
% text(locator(1),"t=s")

